---
title: "Predictors NY"
format:
  html:
    self-contained: true
    embed-resources: true
    toc: true          # optional: adds a table of contents
    theme: cosmo       # optional: Bootstrap theme
    code-fold: show    # optional: collapsible code blocks
    code-tools: true   # optional: adds copy/paste buttons
    toc-depth: 2  
editor: 
  markdown: 
    wrap: 72
---

In the following we match external data to our species.
1. Avonet traits [1]
2. Phylogenetic distinctiveness [2,3]
3. IUCN threat status[4]
4. Range geometry
5. Spatial autocorrelation [5]
6. Lacunarity
7. Global range size [6]

Finally, we merge all predictors together.

```{r}
#| warning: false
#| message: false
#| error: false

rm(list=ls())
gc()
suppressPackageStartupMessages({
library(here)
library(dplyr)
library(skimr)
library(readxl)
library(tictoc)

# Phylo distinct
library(ape)
library(phyloregion)

# IUCN
library(taxize)

# Range geometry
library(sf)
library(tidyverse)
library(purrr)
library(terra)
library(broom)
library(geosphere)

# Spatial autocorrelation

library(spdep)
library(rstatix)
# install.packages("spdep", repos = "https://cloud.r-project.org/") # if not installed


})
```

### Paths

```{r}
#| warning: false
#| message: false
#| error: false

data_sf_path <- 
  here("Demo_NewYork/Data/output/1_data_sf_ny.rds")

sp_data_path <- 
  here("Demo_NewYork/Data/output/1_data_filtered_ny.rds")

tax_path <-
  here("Demo_NewYork/Data/input/Tax_lookup_ny.csv")

avonet_path <- 
  here("Demo_NewYork/Data/input/AVONET Supplementary dataset 1.xlsx")

tree_path <- 
   here("Demo_NewYork/Data/input/MRC_consensus_BirdTree.tre")
```

### Avonet traits

-   Download AVONET Supplementary dataset 1.xlsx here ([[1] Tobias et al. (2022)](https://figshare.com/s/b990722d72a26b5bfead?file=34480856))

-   We placed it here: `Demo_NewYork/Data/input/AVONET Supplementary dataset 1.xlsx` (see path above)


```{r}
#----------------------------------------------------------#
# Load data  -----
#----------------------------------------------------------#

# filtered species list
sci_name <-
  readRDS(sp_data_path) %>%
  distinct(verbatimIdentification, scientificName) # BirdLife 2024 taxonomy

# taxonomic crosswalk with BirdLife 2018 taxonomy as used in Avonet
Tax <- 
  read.csv2(tax_path) %>%
  select(verbatimIdentification, scientificName, ScientificName2018)%>%
  filter(verbatimIdentification %in% sci_name$verbatimIdentification) %>%
  distinct(verbatimIdentification, scientificName, .keep_all=TRUE)

# Extract traits from Avonet
traits <-
  readxl::read_excel(avonet_path, sheet = "AVONET1_BirdLife")  %>%
  dplyr::select(
    Species1, Mass,
    Habitat, Migration, Primary.Lifestyle) %>%
  mutate(
    across(c(Habitat, Migration, Primary.Lifestyle),
           as.factor)) 

# Merge with species names
traits_merged <- traits %>%
  rename("ScientificName2018" = "Species1") %>%
  right_join(Tax) %>%
  select(-ScientificName2018) %>%
  distinct(verbatimIdentification, scientificName, .keep_all = TRUE)

# checks:
traits_merged %>% skimr::skim() # 3 missing


rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","tree_path","sci_name","traits_merged")))
```

### Phylogenetic distinctiveness

This requires the BirdTree [2] tree file, which can be downloaded from [here](https://birdtree.org/). We use the consensus tree that we created based on recommendations from [3]. 
We placed the tree file here: `Demo_NewYork/Data/input/MRC_consensus_BirdTree.tre` (see path above).

```{r}

Tax <- read.csv2(tax_path)[2:7] %>%
  filter(verbatimIdentification %in% sci_name$verbatimIdentification)
tree <- read.tree(tree_path)


# pd for all species in the tree
pd <-
  phyloregion::evol_distinct(
  tree,
  type = "fair.proportion",
  scale = FALSE,
  use.branch.lengths = TRUE) %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "tip.label") %>%
  rename(pd = ".") %>%
  right_join(Tax) %>%
  select(verbatimIdentification, scientificName, pd) %>%
  filter(verbatimIdentification %in% sci_name$verbatimIdentification) %>%
  distinct(scientificName, verbatimIdentification, .keep_all=T)

# checks
colSums(is.na(pd)) # 3 species not matched
pd %>% 
  filter(is.na(pd)) %>% 
  pull(verbatimIdentification)

hist(pd$pd, breaks = 30, 
  main = "Phylogenetic Distinctiveness distribution", 
  xlab = "Phylogenetic Distinctiveness")


rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","sci_name","traits_merged", "pd")))
```

### IUCN Threat status

For this part we need a personal API key to access the IUCN Red List API. You can get one from [here](https://api.iucnredlist.org/). There you have to sign up with an account and request an API token. Once you have it, you can either set it as an environment variable in an .Renviron file e.g., `IUCN_REDLIST_KEY = [key]` or paste it directly into the code below.

- [4] IUCN (2025)

```{r}
#| warning: false
#| message: false
#| error: false

# Download status via Red List API (acces through taxize)
tictoc::tic()

IUCN.list <-
  suppressMessages(iucn_summary(unique(sci_name$scientificName),
    distr.detail = F,
    key =" DwikYETjk5iHVudwZdUDhCYVT5fqFjScBnkE"
    #key = Sys.getenv("IUCN_REDLIST_KEY") # Replace with your API KEY
  ))
tictoc::toc() # 179.11 sec elapsed



# Extract codes and bind to dataframe
IUCN <- 
  lapply(names(IUCN.list), function(name) {
  item <- IUCN.list[[name]]

  # Check if the element is a list and contains `red_list_category`
  if (is.list(item) && 
      !is.null(item$red_list_category) && 
      !is.null(item$red_list_category$code)) {
    
    return(
      data.frame(
        name = name, 
        code = item$red_list_category$code, 
        stringsAsFactors = FALSE)
      )
    
  } else {
    return(
      data.frame(
        name = name, 
        code = NA, 
        stringsAsFactors = FALSE)
      ) # Fill missing with NA
  }
}) %>% 
  do.call(rbind, .) # bind rows to dataframe

# Check NA species: 10 species not found
species_with_NA <-
  IUCN %>% 
  filter(is.na(code))

# try again with those that were not found:
IUCN.list2 <-
  suppressMessages(iucn_summary(unique(species_with_NA$name),
    distr.detail = F,
    key =" DwikYETjk5iHVudwZdUDhCYVT5fqFjScBnkE"
    #key = Sys.getenv("IUCN_REDLIST_KEY")
  ))
# 3 species not found of which 2 are hybrids and Nannopterium auritus which we will look up manually and assign the status by hand

IUCN2 <- lapply(names(IUCN.list2), function(name) {
  item <- IUCN.list2[[name]]

  # Check if the element is a list and contains `red_list_category`
  if (is.list(item) && 
      !is.null(item$red_list_category) && 
      !is.null(item$red_list_category$code)) {
    return(
      data.frame(
        name = name, 
        code = item$red_list_category$code, #
        stringsAsFactors = FALSE)
      )
  } else {
    return(
      data.frame(
        name = name, 
        code = NA, 
        stringsAsFactors = FALSE)
      ) # Fill missing with NA
  }
}) %>% 
  do.call(rbind, .)


# Merge Results 1 and Results 2
IUCN_merged <- 
  rbind(IUCN %>% na.omit(), IUCN2) %>%
  unique() %>%
  # assign code to Nannopterum auritus by hand - because it is on IUCN.
  mutate(code = case_when(name == "Nannopterum auritus" ~ "LC",
    .default = code
  ))

# checks: only the two hybrids left
IUCN_merged %>% 
  filter(is.na(code))

# Reshape
IUCN_df <-
  as.data.frame(IUCN_merged,
    row.names = c(IUCN_merged$name)
  ) %>%
  tibble::rownames_to_column(var = "scientificName") %>%
  right_join(sci_name) %>%
  distinct(verbatimIdentification, scientificName, code)

table(IUCN_df$code)

rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","sci_name","traits_merged", "pd", "IUCN_df")))
```

### Range geometry 

```{r}
# Functions:
## Normalized circularity function (written by Gabriel Ortega Solís)
circNorm <- function(x){
    perimeter <- perim(x)
    area <- expanse(x)
    res <- (perimeter^2) / (4 * pi * area)
    return(res)
}

## Smallest distance from the range centroid to the atlas border
minDist <- function(centroid, borders){
  res <- min(
    distance(
      crds(centroid), 
      crds(borders), 
      lonlat = F)
    )
  return(res)
}

# s2 settings:
sf_use_s2(FALSE) # we use planar projections (i.e., in metres) so this can be turned off

#----------------------------------------------------------#
# Load data -----
#----------------------------------------------------------#

data_sf <-
  readRDS(data_sf_path) %>%
  filter(scalingID == 1 & 
           cells_keep == 1) 

#----------------------------------------------------------#
# Atlas geometries  -----
#----------------------------------------------------------#

sf_current_atlas <- 
  data_sf %>% 
  filter(samplingPeriodID == 1) %>%
  select(geometry) %>%
  st_as_sf() %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_transform(crs = unique(data_sf$verbatimFootprintSRS)) %>%
  st_make_valid()

plot(sf_current_atlas)

# vectorize atlas
terra_current_atlas <- vect(sf_current_atlas)

plot(terra_current_atlas)

# Store borders as lines
atlas_border_lines <- terra::as.lines(terra_current_atlas)

#----------------------------------------------------------#
# Single species range geometries  -----
#----------------------------------------------------------#

grouped_list <- 
  data_sf %>% 
  group_by(samplingPeriodID, verbatimIdentification) %>%
  filter(species_keep == 1) %>%
  group_split()

tictoc::tic()
range_geometries <- map_dfr(grouped_list, function(dta) {
  
  sp_range <- 
    dta %>%
      dplyr::select(geometry, siteID, datasetID) %>%
      st_transform(crs = st_crs(atlas_border_lines)) %>%
      summarise() %>%
      terra::vect()
  
  sp_range_centroid <- 
    centroids(sp_range)
  
  # checks
  # plot(sp_range); plot(sp_range_centroid, add = T, col = "red")
  
  # Compute range attributes
  tibble(
    datasetID = unique(dta$datasetID),
    samplingPeriodID = unique(dta$samplingPeriodID),
    verbatimIdentification = unique(dta$verbatimIdentification),
    circNorm = circNorm(sp_range), 
    minDist_toBorder_centr = minDist(sp_range_centroid, atlas_border_lines)
    )
  
})
tictoc::toc() # 108.94 sec elapsed

# checks:
range_geometries %>% 
  skimr::skim()


rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","sci_name","traits_merged", "pd", "IUCN_df", "range_geometries")))
```

### Spatial autocorrelation
We calculate the difference in expected and observed join counts for presences across the atlas for each species and sampling period. This is done using the `joincount.test()` function from the `spdep` package. The join count test is a measure of spatial autocorrelation that counts the number of pairs of neighboring sites that have the same value (in this case, presence or absence of a species). In the calculation of delat join count, we follow still unpublished work by Soria et al. (2025) [5]

```{r}
# CRS
ny_crs <-"epsg:32118"  #verbatimFootprintSRS (NAD83 / New York Long Island)

# Read data
data_sf <- 
  readRDS(data_sf_path) %>%
  filter(scalingID == 1 &
           cells_keep == 1 &
           species_keep == 1) %>%
  st_transform(crs = ny_crs)


# calculate join count across samplingPeriods for each species
sampling_periods <- c(1,2)

# Safe versions of spatial autocorrelation functions
safe_joincount_test <- possibly(function(x, nb) joincount.test(as.factor(x), nb2listw(nb, style = "B", zero.policy = TRUE)), otherwise = NA)

tictoc::tic()
joincount_df <- map_dfr(sampling_periods, function(sampling_period){
  tp_subset <- 
    data_sf %>%
    filter(samplingPeriodID == sampling_period)
  
  sp_names <- 
    tp_subset %>% 
    st_drop_geometry() %>% 
    distinct(verbatimIdentification) %>% 
    pull() 
  
  all_sites <- 
    data_sf %>%
    select(siteID, geometry) %>%
    unique()
  
  map_dfr(sp_names, function(sp_name){
    
    sp_data <- 
      tp_subset %>%
      st_drop_geometry() %>%
      filter(verbatimIdentification == sp_name) %>%
      full_join(all_sites, by = "siteID") %>%
      mutate(
        presence = if_else(is.na(verbatimIdentification), 0, 1),
        verbatimIdentification = sp_name) %>%
      st_as_sf()
    
    # checks:
    # plot(sp_data["presence"])
    
    # Count the number of presences
    num_pres <- sum(sp_data$presence, na.rm = TRUE)
    
    # Creating neighbors (queen)
    nb_q <- spdep::poly2nb(sp_data, queen = TRUE)

    # Join count test
    jc_res_q <- safe_joincount_test(sp_data$presence, nb_q)
    
    # Save results in a dataframe
    tibble(
      samplingPeriodID = sampling_period,
      verbatimIdentification = sp_name,
      presence_n = num_pres,
      joincount_statistic = ifelse(is.na(jc_res_q), NA,
                                   jc_res_q[[2]]$estimate[1])[[1]],
      joincount_expectation = ifelse(is.na(jc_res_q), NA,
                                     jc_res_q[[2]]$estimate[2])[[1]],
      joincount_p_val = ifelse(is.na(jc_res_q), NA,
                               jc_res_q[[2]]$p.value)[[1]],
      joincount_delta = ifelse(is.na(jc_res_q), NA,
                               ((jc_res_q[[2]]$estimate[1] /
                                   num_pres) -
                                  (jc_res_q[[2]]$estimate[2] /
                                     num_pres)))[[1]]
      ) %>%
      unique()

    
  })
  
})

tictoc::toc() # 1376.94 sec elapsed

# checks:
joincount_df %>%
  skimr::skim()

rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","sci_name","traits_merged", "pd", "IUCN_df", "range_geometries", "joincount_df")))
```


### Lacunarity

```{r}
# Load required packages
package_list <-
  c("here", "sf", "tidyverse", "purrr", "tictoc", "terra")
x <-
  lapply(package_list, require, character.only = TRUE)
rm(x)

# turn spherical geometry off (we are working with planar proj. in meters)
sf_use_s2(FALSE) # we use data in meters

# Original CRS
ny_crs <-"epsg:32118"  #verbatimFootprintSRS (NAD83 / New York Long Island)

# Read in sf data
data_sf <- 
  readRDS(data_sf_path) %>% 
  filter(scalingID == 1) %>%
  st_transform(crs = ny_crs)
```


```{r}
# Prepare country template for rasterizing ----
resolution <- c(5000, 5000) # in metres

bbox_ny <- 
  data_sf %>% 
  select(siteID, geometry) %>%
  unique() %>% 
  st_bbox()

country_border <-   
  data_sf %>% 
  select(siteID, geometry) %>%
  unique() %>%
  st_union() %>%
  vect()

# create masked template for rasterizing based on NY outlines
template <- 
  rast(
    ext(bbox_ny$xmin, bbox_ny$xmax, bbox_ny$ymin, bbox_ny$ymax),
    resolution = resolution,
    crs = ny_crs)
values(template) <- 0
masked_template <- mask(template, country_border)

# check:
plot(masked_template)
```


```{r}
# Prepare species data ----

unsampled_sites_expanded <- 
  data_sf %>%
  filter(cell_sampling_repeats == 0) %>%
  distinct(siteID, geometry) %>%
  tidyr::expand_grid(samplingPeriodID = c(1,2)) %>%
  mutate(verbatimIdentification = NA)

# combine with sampled cells
data_sf_with_unsampled <- 
  bind_rows(
    data_sf %>%
      filter(cell_sampling_repeats == 2 | !is.na(verbatimIdentification)),
    unsampled_sites_expanded
  )

# get filtered species list
species_list <- 
  sci_name %>% 
  pull(verbatimIdentification) %>% 
  unique()

# rasterize ranges

raster_list <- replicate(2, list())
for (tp in seq_along(c(1,2))){
  
  raster_list[[tp]] <- list()
  
  data_tp_i <- 
    data_sf_with_unsampled %>%
    filter(samplingPeriodID == tp)
  
  for (sp_name in species_list){
    
    this_data <- 
      data_tp_i %>%
      filter(verbatimIdentification == sp_name & samplingPeriodID == tp) %>%
      st_as_sf() %>%
      st_transform(crs = ny_crs) %>%
      mutate(
        presence = 1)
  
  r <- 
    rasterize(this_data, 
              masked_template, 
              field = "presence", 
              update = TRUE)
    
  raster_list[[tp]][[sp_name]] <- r
    
  }
}

# example plot
plot(r)
  
rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","sci_name","traits_merged", "pd", "IUCN_df", "range_geometries", "joincount_df", "raster_list")))
gc()

```

```{r}
# Calculate lacunarity across sub-lists
#----------------------------------------------------------#
# Function to run terra::focal() for multiple window-sizes -----
#----------------------------------------------------------#
apply_focal_multiple_w <- function(r, w_vec) {

  results <- lapply(w_vec, function(w) {
    f <- focal(r, w = w, fun = sum, na.policy="omit", na.rm=TRUE)

    # n_S_r = number of filled cells in the box (of size r)
    # Q_S_r = normalized frequency distribution of S for box size r

    box_masses_int <- as.integer(values(f, na.rm = T))
    max_value <- max(box_masses_int, na.rm = TRUE)

    # initialize empty vector (add +1 to enable indexing with 0 values)
    n_S_r <- integer(max_value + 1L)
    for (val in box_masses_int) {
      if (val >= 0 && val <= max_value) {
        n_S_r[val + 1L] <- n_S_r[val + 1L] + 1
      }
    }
    Q_S_r <- n_S_r / length(box_masses_int[!is.na(box_masses_int)])
    S_vals <- seq_along(Q_S_r) - 1
    Z_1 <- sum(S_vals * Q_S_r) #first moment = mean
    Z_2 <- sum((S_vals^2) * Q_S_r) # second moment = variance (mean of squared values)

    if (Z_1 == 0) {
      return(NA_real_)
    }
    lac <- Z_2 / (Z_1 * Z_1)

    return(lac)
  })


  names(results) <- paste0("w=", w_vec)
 r_name <- sp_name
  message(r_name, ": done")
  # lac_res <- data.frame(lac = do.call(rbind, results), w = w_vec, species = r_name)
  # return(lac_res)


  this_out <- dplyr::tibble(name = rep(r_name, length(results)),
                            r = w_vec, "ln(r)" = log(w_vec),
                            Lac = results %>% do.call(rbind,.) %>% .[,1], "ln(Lac)" = log(results %>% do.call(rbind,.) %>% .[,1]))
  return(this_out)
}
```

```{r}
#----------------------------------------------------------#
# Run focal function across multiple window-sizes -----
#----------------------------------------------------------#

w_vec <- c(3L, 5L, 9L, 17L, 33L) # window sizes

tictoc::tic()
res_list <- replicate(2, list())

for (tp_i in seq_along(raster_list)){
  res_list[[tp_i]] <- list()
  for(sp_i in seq_along(raster_list[[tp_i]])){
    
    sp_name <- names(raster_list[[tp_i]])[sp_i]
    r <- raster_list[[tp_i]][[sp_i]]
    test_res <- apply_focal_multiple_w(r, w_vec)
    
    # add species info
    test_res$samplingPeriodID <- tp_i
    test_res$verbatimIdentification <- sp_name
    
    
    res_list[[tp_i]][[sp_i]] <- test_res
  }
}
tictoc::toc() # 96.32 sec elapsed

df_res1 <- do.call(rbind, res_list[[1]])
df_res2 <- do.call(rbind, res_list[[2]])

lac_results <- 
  rbind(df_res1, df_res2) %>% 
  select(-name) %>%
  as.data.frame()

names(lac_results) <- c("r", "ln(r)", "lac", "ln(lac)", "samplingPeriodID", "verbatimIdentification")


# Calculate mean across increasing window sizes
mean_lac <- lac_results %>%
  ungroup() %>%
  group_by(samplingPeriodID, verbatimIdentification) %>%
  summarize(mean_lnLac = mean(`ln(lac)`, na.rm = TRUE), .groups = "keep") 


hist(mean_lac$mean_lnLac, breaks = 30, 
     main = "Lacunarity distribution", 
     xlab = "Mean Lacunarity (ln)")

rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","sci_name","traits_merged", "pd", "IUCN_df", "range_geometries", "joincount_df", "mean_lac")))
gc()

```

### Global range size
We again use the BirdLife range polygons [6], which can be downloaded from [here](http://datazone.birdlife.org/species/requestdis). We use the Mollweide projection to calculate the global range size in km². The BirdLife shapefile is in WGS84, so we transform it to Mollweide before calculating the area.

```{r}
#| eval: true
#| include: true

# planar CRS
Mollweide_CRS <- 
'PROJCS["ProjWiz_Custom_Mollweide",
 GEOGCS["GCS_WGS_1984",
  DATUM["D_WGS_1984",
   SPHEROID["WGS_1984",6378137.0,298.257223563]],
  PRIMEM["Greenwich",0.0],
  UNIT["Degree",0.0174532925199433]],
 PROJECTION["Mollweide"],
 PARAMETER["False_Easting",0.0],
 PARAMETER["False_Northing",0.0],
 PARAMETER["Central_Meridian",0],
 UNIT["Meter",1.0]]'

sf_use_s2(FALSE)

# Taxonomic match table
Tax <- 
  sci_name # Taxa

# Global ranges
tictoc::tic()
BirdLife_split <- 
  st_read(here("Data/input/shp_global/BirdLife.shp"), quiet = TRUE)%>% 
  filter(sci_name %in% c(Tax$scientificName)) %>%
  st_transform(crs = Mollweide_CRS)%>%
  group_by(sci_name) %>%
  st_make_valid() %>% # takes a while...
  group_by(sci_name) %>% 
  group_split()
  

# loop through list and sum up the combined range areas for each species
range_size_df <- 
  map_dfr(BirdLife_split, function(this_species){
    
    message(paste0(unique(this_species$sci_name), "... done!"))
  
    this_species %>%
      group_by(sci_name) %>%
      summarise(.groups = "keep") %>%
      # Convert to km²
      group_by(sci_name) %>%
      mutate(GlobRangeSize_km2 = as.numeric(st_area(.))/ 1e6) %>%  
      st_drop_geometry()%>%
      rename(scientificName = sci_name)
  
    
    })
tictoc::toc()
 
# Merge with verbatimIdentification
range_size_df2 <- 
  range_size_df %>%
  right_join(Tax, by = "scientificName") %>%
  distinct(scientificName, verbatimIdentification, .keep_all = TRUE) 

# checks
colSums(is.na(range_size_df2))
range(range_size_df2$GlobRangeSize_km2, na.rm=TRUE)
hist(log(range_size_df2$GlobRangeSize_km2))



rm(list = setdiff(ls(),c("data_sf_path","tax_path","sp_data_path","sci_name","traits_merged", "pd", "IUCN_df", "range_geometries", "joincount_df", "mean_lac", "range_size_df2")))
gc()
```

## Merge Predictors together
```{r}
# Set 1: External data
predictors_1 <- 
  full_join(range_size_df2, traits_merged) %>%
  full_join(IUCN_df %>% mutate(code = as.factor(code)) %>% rename("IUCN" = "code")) %>%
  full_join(pd) %>%
  distinct(verbatimIdentification, .keep_all = TRUE)

colSums(is.na(predictors_1))
glimpse(predictors_1)

# Set 2: Calculated from atlas

big_table <- 
  readRDS(here("Demo_NewYork/Data/output/A_Big_table_ny.rds"))

predictors_2 <- 
  full_join(big_table, joincount_df) %>%
  full_join(range_geometries) %>%
  full_join(mean_lac) %>%
  distinct(samplingPeriodID, verbatimIdentification, .keep_all = TRUE)

colSums(is.na(predictors_2))
glimpse(predictors_2)
```

#### Modify predictors for modeling 
```{r}
# Merge together
data_final <- 
  full_join(predictors_2, predictors_1) %>%
  filter(samplingPeriodID == 1) %>%
  mutate(
    across(
      where(is.character) & 
        !matches("verbatimIdentification") & 
        !matches("scientificName"),
      as.factor)) %>%
  mutate(
    across(
      c("Habitat", "IUCN", "Migration", "Primary.Lifestyle"),
      as.factor)) %>%
  distinct(verbatimIdentification, .keep_all = TRUE) %>%
  ungroup() %>%
  mutate(
    Habitat_5 = 
      as.factor(
        case_when(
          Habitat %in% c("Grassland", "Shrubland", 
                         "Desert", "Rock") ~ "open",
          Habitat %in% c("Woodland", "Forest") ~ "closed",
          Habitat %in% c("Coastal", "Marine") ~ "marine",
          Habitat %in% c("Wetland", "Riverine") ~ "freshwater",
          Habitat == "Human Modified" ~ "human",
          .default = NA_character_)
        )
    ) %>%
  mutate(
    Generalism = 
      as.factor(
        case_when(
          Primary.Lifestyle == "Generalist" ~ 1,
          .default = 0)
        )
    ) %>%
  mutate(
    Threatened = 
      as.factor(
        case_when(
          IUCN %in% c("LC") ~ 0,
          is.na(IUCN) ~ NA,
          .default = 1)
        )
    ) %>%
  select(
    verbatimIdentification, scientificName, 
    log_R2_1, log_R2_1_per_year, Jaccard_dissim, 
    D_AOO_a, AOO, joincount_delta, mean_lnLac,
    circNorm, minDist_toBorder_centr, 
    GlobRangeSize_km2, Mass, Migration, 
    pd, Habitat_5, Generalism, Threatened
  )


data_final %>% 
  skimr::skim()


write.csv(data_final, here::here("Demo_NewYork/Data/output/data_final_ny.csv"))

```

## References

[1] Tobias, J. A., Sheard, C., Pigot, A. L., Devenish, A. J., Yang, J., Sayol, F., Neate-Clegg, M. H., Alioravainen, N., Weeks, T. L., Barber, R. A., & others. (2022). AVONET: morphological, ecological and geographical data for all birds. Ecology Letters, 25(3), 581–597.


[2] Jetz, W., Thomas, G. H., Joy, J. B., Hartmann, K., & Mooers, A. O. (2012). The global diversity of birds in space and time. Nature, 491(7424), 444–448.

[3] Rubolini, D., Liker, A., Garamszegi, L. Z., Møller, A. P., & Saino, N. (2015). Using the BirdTree.org website to obtain robust phylogenies for avian comparative studies: A primer. Current Zoology, 61(6), 959–965. https://doi.org/10.1093/czoolo/61.6.959

[4] IUCN (2025). The IUCN Red List of Threatened Species. Version 2025-1. https://www.iucnredlist.org. Accessed 14 June 2025.

[5] Soria, C. D., Ortega-Solís, G. R., Wölke, F. J. R., Barták, V., Tschernosterová, K., Bejček, V., Herrando, S., Mikuláš, I., Šťastný, K., Ueta, M., Voříšek, P., & Keil, P. (in preparation). Spatial autocorrelation of species diversity and distributions in time and across spatial scales.

[6] BirdLife International & Handbook of the Birds of the World. (2024). Bird species distribution maps of the world (Version 9.0) [Dataset]. http://datazone.birdlife.org/species/requestdis

