---
title: "Simulating J dissimilarity under different scenarios of scale, occupancy, temporal change magnitude"
author: "Friederike Wölke"
format: html
editor: visual
---

# Simulating *J dissimilarity* under different scenarios of scale, occupancy, temporal change magnitude

## ℹ️Parameters to simulate

-- universe size\
– occu 1\
– flip type: directional/random\
– number of changes\
– number of runs

## ❇️ Simulation function

```{r}
#| eval: false
#| include: true

my_jaccard_abc <- function(set1, set2){
  a <- length(intersect(set1, set2))
  b <- length(setdiff(set2, set1))
  c <- length(setdiff(set1, set2))
  res <- data.frame(j = a / (a + b + c),
             intersect = a,
             unique_1 = c,
             unique2 = b)
  return(res)
}



set.seed(123) # For reproducibility

simulate_jaccard <- function(lengths, occu1, flip_types, n_runs = 1) {
  # Expand grid for parameter combinations
  params <- expand.grid(lengths, occu1, flip_types, stringsAsFactors = FALSE)
  colnames(params) <- c("length", "occu1", "flip_type")

  # Initialize storage
  jaccard_results <- list()

  # Loop through parameter combinations
  for (row_i in seq_len(nrow(params))) {
    length_i <- params$length[row_i]
    occu_1_i <- params$occu1[row_i]
    flip_type <- params$flip_type[row_i]

    # Ensure the number of occupied sites does not exceed the vector length
    occu_1_i <- min(occu_1_i, length_i)

    # Repeat the simulation `n_runs` times
    for (run in seq_len(n_runs)) {
      # Generate set1 (initially occupied sites)
      set1_i <- sample(1:length_i, occu_1_i, replace = FALSE)

      # Sequence of number of changes (from 1 to max_vector_length)
      n_changes <- seq(from = 1, to = length_i, by = 1)

      # Dataframe to store results for this run
      jaccard_temp <- data.frame(num_changes = n_changes, jaccard = NA, run = run)

      set2_temp_list <- vector("list", length(n_changes))

      # Modify set1 stepwise and store set2 states
      for (n_i in seq_along(n_changes)) {
        set2_temp <- set1_i
        changes_i <- n_changes[n_i]

        if (flip_type == "random") {
          flip_sites <- sample(1:length_i, size = changes_i, replace = FALSE)

          set2_i <- union( # Combine left overs after 1 to 0 flips with new cells from 0 to 1 flips
            setdiff(set2_temp, flip_sites), # Remove indices in `flip_sites` already in `set2_temp`
            intersect(flip_sites, setdiff(1:length_i, set2_temp)) # Add indices not in `set2_temp` (0 to 1; colonizations)
          )
        } else if (flip_type == "from_1_to_0") {
          sample_mat <- if (length(set1_i) == 1) rep(set1_i, 2) else set1_i
          flip_sites <- sample(x = sample_mat, size = min(changes_i, length(set1_i)), replace = FALSE)
          set2_i <- setdiff(set2_temp, flip_sites) # Remove flipped sites from set1
          set2_i <- if (length(set2_i) == 0) NA else set2_i
        }

        # Store modified set2
        set2_temp_list[[n_i]] <- set2_i

        # Compute Jaccard similarity
        jaccard_res <- my_jaccard_abc(set1_i, set2_i)

        jaccard_temp$jaccard[n_i] <- jaccard_res$j
        jaccard_temp$intersect[n_i] <- jaccard_res$intersect
        jaccard_temp$unique_1[n_i] <- jaccard_res$unique_1
        jaccard_temp$unique_2[n_i] <- jaccard_res$unique_2
        jaccard_temp$N_occu_1[n_i] <- jaccard_res$N_occu_1
        jaccard_temp$N_occu_2[n_i] <- jaccard_res$N_occu_2
        jaccard_temp$length[n_i] <- length_i
        jaccard_temp$occu1[n_i] <- occu_1_i
        jaccard_temp$flip_type[n_i] <- flip_type
      }

      # Append results for this run
      jaccard_results[[length(jaccard_results) + 1]] <- jaccard_temp
    }
  }

  # Combine Jaccard results into a single dataframe for visualization
  jaccard_df <- do.call(rbind, jaccard_results)
  return(jaccard_df)
}

```

## 🏎️Run simulations

```{r}
#| eval: false
#| include: true

# Example usage
lengths <- c(10, 100, 1000)
occu1 <- c(2^(0:floor(log2(1000))), 999)
flip_types <- c("random", "from_1_to_0")
n_runs <- 100 # Repeat each parameter combination 100 times



jaccard_df <- 
  simulate_jaccard(lengths, occu1, flip_types, n_runs)

saveRDS(jaccard_df, "Code/Sensitivity/Jaccard_Prevalence_Simulations/res/jaccards_sim_absolute_100runs.rds")

```

## 🔍Inspect results

```{r}
#| label: libraries
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(skimr))

```

```{r}
#| label: read-data
jaccard_df <- 
  readRDS(
    here("Data/output/temp", 
         "C_02_Jaccard_simulations_100.rds")) %>% 
  unique()
```

```{r}
#| label: process-data
# quick check
str(jaccard_df)


# organize data frame to inspect
sim_df <- data.frame(
  run = jaccard_df$run,
  universe_size = jaccard_df$length,
  change_type = as.factor(jaccard_df$flip_type),
  n_changes = jaccard_df$num_changes,
  j_dissim = 1-round(jaccard_df$jaccard, 4),
  AOO_1 = jaccard_df$N_occu_1,
  AOO_2 = jaccard_df$N_occu_2,
  a = jaccard_df$intersect,
  b = jaccard_df$unique_2,
  c = jaccard_df$unique_1,
  d = jaccard_df$length - jaccard_df$intersect - jaccard_df$unique_2 - jaccard_df$unique_1
) %>%
  ungroup() %>%
  filter(AOO_1 != 0 & AOO_2 != 0) %>%
  filter(AOO_1 != universe_size) %>%
  filter(n_changes != universe_size) %>%
  mutate(
    log_ratio = log(AOO_2/AOO_1),
    rel_persistance = a/universe_size,
    change_intensity = (b+c)/AOO_1) %>%
  ungroup() %>%
  mutate(
    trend = case_when(b > c ~ "expanding range",
                      b < c ~ "declining range",
                      b == c ~ "A_net zero trend", #added A to sort levels for regression
                      .default = NA))

sim_df %>% filter(is.na(trend)) # no NAs
```

```{r}
# create expanding trend for directional simulations from declining

expanding <- 
  sim_df %>% 
  filter(change_type == "from_1_to_0") %>%
  mutate(change_type = "from_0_to_1") %>%
  mutate(log_ratio = log(AOO_1/AOO_2)) %>% 
  rename("b" = "c", 
         "c" = "b", 
         "AOO_1" = "AOO_2", 
         "AOO_2" = "AOO_1") %>%
  mutate(
    trend = case_when(b > c ~ "expanding range",
                      b < c ~ "declining range",
                      b == c ~ "A_net zero trend", 
                      #added A to sort levels for regression
                      .default = NA))

expanding %>% filter(is.na(trend)) # no NAs  
  
sim_df <- 
  sim_df %>% 
  full_join(expanding) %>%
  mutate(trend = as.factor(trend)) %>%
  mutate(change_type2 = case_when(change_type == "random" ~ "random", change_type %in% c("from_1_to_0", "from_0_to_1") ~ "directional",
                                  .default = NA))


sim_df %>% 
  filter(change_type == "from_0_to_1") %>%
  group_by(trend) %>% 
  skim(log_ratio, a,b,c)

```

```{r}
#| label: j-summary-1

# quick check:
sim_df %>%
  filter(change_type == "random") %>%
  group_by(universe_size, AOO_1, n_changes, change_type, b, c) %>%
  summarize(mean_j = mean(j_dissim),
            sd_j = sd(j_dissim),
            n = n())


sim_df %>% 
  filter(change_type == "random" & universe_size == 1000) %>%
  group_by(trend) %>% 
  skimr::skim()

# --> for directional change, there is no variation in J.
```

## random x large (10 simulations)

```{r}
#| label: data-subset-random-and-large
random_large <- 
  sim_df %>% 
  filter(universe_size == 1000) %>% 
  filter(change_type == "random") %>% 
  #filter(run <= 10) %>%
  select(-universe_size, -change_type)
```

#### Correlation analysis

```{r}
#| label: correlations
library(corrr)

# Select numeric columns and compute correlation with targets
cor_df <- random_large %>%
  select(where(is.numeric)) %>%
  correlate() %>%
  focus(j_dissim, log_ratio) %>%
  mutate(j_dissim = round(j_dissim, 2),
         log_ratio = round(log_ratio, 2))

print(cor_df)

random_large %>%
  select(where(is.numeric)) %>%
  correlate() %>%
  autoplot()
```

#### Regression models

```{r}
#| label: regression-models
# Add interaction term
random_large <- random_large %>%
  mutate(AOO_1_n_changes = AOO_1 * n_changes)

# j_dissim model
model_jaccard <- lm(j_dissim ~  AOO_1:n_changes + a + b + c + d + trend, data = random_large)
summary(model_jaccard)

# log_ratio model
model_log_ratio <- lm(log_ratio ~ AOO_1:n_changes + a + b + c + d + trend, data = random_large)
summary(model_log_ratio)

model_log_ratio$coefficients %>% round(2)


```

```{r}
#| fig.height: 12
#| fig.width: 10
#| label: j-by-net-area-trend

# Boxplot for j_dissim
ggplot(sim_df, aes(x = trend, y = j_dissim, fill = trend)) +
  geom_hline(yintercept = 0.5, col = "darkgrey")+
  geom_boxplot(outlier.colour = "darkgrey") +
  ggthemes::theme_par() +
  labs(title = "Jaccard Dissimilarity by Trend", x = "Trend", y = "j_dissim")+
  facet_grid(universe_size ~ change_type2)+
  scale_fill_manual(values = c("#ffffbf", "#d7191c", "#1a9641"))+
    theme(
    axis.text.x = element_text(
      angle = 45, 
      vjust = 0.9, 
      hjust = 1))

ggsave(filename = "J_simulations_j_trends_boxplots.bmp", path = here::here("Figures/C_validation/"), width = 10, height = 12)

# Boxplot for log_ratio
ggplot(sim_df, aes(x = trend, y = log_ratio, fill = trend)) +
  geom_hline(yintercept = 0, col = "darkgrey")+
  geom_boxplot(outlier.colour = "darkgrey") +
  ggthemes::theme_par() +
  labs(title = "Log Ratio by Trend", x = "Trend", y = "log_ratio")+
  facet_grid(universe_size ~ change_type2)+
  scale_fill_manual(values = c("#ffffbf", "#d7191c", "#1a9641"))+
  theme(
    axis.text.x = element_text(
      angle = 45, 
      vjust = 0.9, 
      hjust = 1))

ggsave(filename = "J_simulations_logRatio_trends_boxplots.svg", path = here::here("Figures/C_validation/"), width = 10, height = 12)
```

```{r}
# what about the large tail for declining and expanding ranges in the directional trial? 

sim_df %>% 
  filter(change_type != "random") %>% 
  filter(j_dissim != 1) %>% 
  ggplot()+
  geom_histogram(aes(j_dissim))


sim_df %>% 
  filter(change_type != "random" & universe_size != 10) %>% 
  filter(j_dissim != 1) %>% 
  group_by(change_type, universe_size) %>% 
  ggplot()+
  geom_point(aes(y = j_dissim, 
                 x = change_intensity, 
                 col = change_intensity))+
  facet_wrap(change_type~universe_size, scales = "free")+
  scale_color_distiller(type = "div")

# Change intensity is:
# the ratio between the number of changed cells and AOO_1:
# change intensity = (b+c)/AOO_1
# where b = colonized cells, c = extirpated cells

# Explanation:
# I.
# ratio approaches 0 when low change 
# - i.e., when the number of lost + gained cells is low compared to AOO_1.
# - b+c < AOO_1
# --> leads to low J dissimilarity


# II.
#  ratio approaches 1 when high change 
# - i.e., when the number of lost + gained cells is comparable to AOO_1 
# - b+c = AOO_1
# --> leads to high J dissimilarity for both directions of change

# III.
# ratio approaches 2 when change is higher than initial occupancy
# - i.e., when more cells are lost + gained than occupied in AOO_1,
# - b+c >> AOO_1
# - --> leads to moderate to high J when range is expanding
# - --> leads to high J if range is declining

# or
# - i.e., when all no cells persist, i.e., all cells from AOO_1 are lost and some are gained (c = AOO_1 & b > 0)
# - a = 0
# --> leads to high J for both directions of change
# or
# - i.e., with strong expansion (b >> AOO_1)
# --> leads to moderate to strong J for both directions
```

```{r}
random_large %>% 
  filter(run <= 20) %>%
  ggplot(aes(x = log_ratio, y = j_dissim, col = trend))+ 
  geom_jitter(cex = 0.1,  alpha = 0.3)+
  ggthemes::theme_few()
```

```{r}
random_large %>%
  ggplot()+
  geom_point(aes(
    y = log_ratio, 
    x = log(AOO_1), 
    col = log(n_changes)), 
    alpha = 0.1)+
  ggthemes::theme_par()+
  facet_wrap(~ trend)

random_large %>%
  ggplot()+
  geom_point(aes(
    y = j_dissim, 
    x = log(AOO_1), 
    col = log(n_changes)), 
    alpha = 0.1)+
  ggthemes::theme_par()+
  facet_wrap(~ trend)


```

```{r}


plot_ly(random_large, 
        z = ~ n_changes, 
        y = ~ j_dissim, 
        x = ~ AOO_1, 
        type = "scatter3d",
        mode = "markers"
        ) %>% 
  add_markers(color = ~ d) %>%
    layout(title = "3D Scatter Plot",
         scene = list(xaxis = list(title = "AOO 1"),
                      yaxis = list(title = "Jaccard"),
                      zaxis = list(title = "n changes")))


```

#### 3D surface plot for Jaccard (GAM smoothed)

```{r}
library(plotly)
# Create a grid for plotting (if needed, you can smooth with a GAM first)
library(mgcv)


gam_mod <- gam(j_dissim ~ te(AOO_1, n_changes), data = random_large)

# Predict over a grid
grid_df <- expand.grid(
  AOO_1 = seq(min(random_large$AOO_1), max(random_large$AOO_1), length = 50),
  n_changes = seq(min(random_large$n_changes), max(random_large$n_changes), length = 50)
)

grid_df$pred_jaccard <- predict(gam_mod, newdata = grid_df)

# Reshape into matrix for surface plot
z_matrix <- matrix(grid_df$pred_jaccard, nrow = 50, ncol = 50)

p <- plot_ly(
  x = ~unique(grid_df$AOO_1),
  y = ~unique(grid_df$n_changes),
  z = ~z_matrix
) %>%
  add_surface(opacity = 0.7) %>%
  layout(
    title = "Jaccard Surface",
    scene = list(
      xaxis = list(title = "Initial Occupancy"),
      yaxis = list(title = "Number of Changes"),
      zaxis = list(title = "Predicted Jaccard")
    )
  )



## add boxes outside of cube
# Compute axis limits
xlim <- range(grid_df$AOO_1)
ylim <- range(grid_df$n_changes)
zlim <- range(grid_df$pred_jaccard)

# Define cube corners
x_vals <- xlim
y_vals <- ylim
z_vals <- zlim

# Function to create lines between pairs of corners
add_box_edges <- function(plot) {
  lines <- list(
    # bottom rectangle
    list(x = x_vals, y = rep(y_vals[1], 2), z = rep(z_vals[1], 2)),
    list(x = x_vals, y = rep(y_vals[2], 2), z = rep(z_vals[1], 2)),
    list(x = rep(x_vals[1], 2), y = y_vals, z = rep(z_vals[1], 2)),
    list(x = rep(x_vals[2], 2), y = y_vals, z = rep(z_vals[1], 2)),

    # top rectangle
    list(x = x_vals, y = rep(y_vals[1], 2), z = rep(z_vals[2], 2)),
    list(x = x_vals, y = rep(y_vals[2], 2), z = rep(z_vals[2], 2)),
    list(x = rep(x_vals[1], 2), y = y_vals, z = rep(z_vals[2], 2)),
    list(x = rep(x_vals[2], 2), y = y_vals, z = rep(z_vals[2], 2)),

    # vertical lines
    list(x = rep(x_vals[1], 2), y = rep(y_vals[1], 2), z = z_vals),
    list(x = rep(x_vals[1], 2), y = rep(y_vals[2], 2), z = z_vals),
    list(x = rep(x_vals[2], 2), y = rep(y_vals[1], 2), z = z_vals),
    list(x = rep(x_vals[2], 2), y = rep(y_vals[2], 2), z = z_vals)
  )

  for (edge in lines) {
    plot <- plot %>% add_trace(
      type = 'scatter3d',
      mode = 'lines',
      x = edge$x,
      y = edge$y,
      z = edge$z,
      line = list(color = 'black', width = 2),
      showlegend = FALSE
    )
  }

  return(plot)
}

# Add the box outlines
p_with_box <- add_box_edges(p)
p_with_box



# Add contours
fig <- p_with_box %>% 
add_surface(opacity = 0.7,
  contours = list(
    z = list(
      show=TRUE,
      usecolormap=TRUE,
      highlightcolor="#ff0000",
      project=list(z=TRUE)
      )
    )
  )
fig %>% layout(
    scene = list(
      camera=list(
        eye = list(x=-0.8, y=0.8, z=0.6)
        )
      )
  )

fig %>% layout(scene = list(camera = list(eye = list(x = -1.25, y = 1.25, z = 1.25))))
```

```{r}
library(plotly)
# Create a grid for plotting (if needed, you can smooth with a GAM first)
library(mgcv)


gam_mod <- gam(log_ratio ~ te(AOO_1, n_changes), data = random_large)

# Predict over a grid
grid_df <- expand.grid(
  AOO_1 = seq(min(random_large$AOO_1), max(random_large$AOO_1), length = 50),
  n_changes = seq(min(random_large$n_changes), max(random_large$n_changes), length = 50)
)

grid_df$pred_lr <- predict(gam_mod, newdata = grid_df)

# Reshape into matrix for surface plot
z_matrix <- matrix(grid_df$pred_lr, nrow = 50, ncol = 50)

p <- plot_ly(
  x = ~unique(grid_df$AOO_1),
  y = ~unique(grid_df$n_changes),
  z = ~z_matrix
) %>%
  add_surface(opacity = 0.7) %>%
  layout(
    title = "Log Ratio Surface",
    scene = list(
      xaxis = list(title = "Initial Occupancy"),
      yaxis = list(title = "Number of Changes"),
      zaxis = list(title = "Predicted log ratio")
    )
  )



## add boxes outside of cube
# Compute axis limits
xlim <- range(grid_df$AOO_1)
ylim <- range(grid_df$n_changes)
zlim <- range(grid_df$pred_lr)

# Define cube corners
x_vals <- xlim
y_vals <- ylim
z_vals <- zlim

# Function to create lines between pairs of corners
add_box_edges <- function(plot) {
  lines <- list(
    # bottom rectangle
    list(x = x_vals, y = rep(y_vals[1], 2), z = rep(z_vals[1], 2)),
    list(x = x_vals, y = rep(y_vals[2], 2), z = rep(z_vals[1], 2)),
    list(x = rep(x_vals[1], 2), y = y_vals, z = rep(z_vals[1], 2)),
    list(x = rep(x_vals[2], 2), y = y_vals, z = rep(z_vals[1], 2)),

    # top rectangle
    list(x = x_vals, y = rep(y_vals[1], 2), z = rep(z_vals[2], 2)),
    list(x = x_vals, y = rep(y_vals[2], 2), z = rep(z_vals[2], 2)),
    list(x = rep(x_vals[1], 2), y = y_vals, z = rep(z_vals[2], 2)),
    list(x = rep(x_vals[2], 2), y = y_vals, z = rep(z_vals[2], 2)),

    # vertical lines
    list(x = rep(x_vals[1], 2), y = rep(y_vals[1], 2), z = z_vals),
    list(x = rep(x_vals[1], 2), y = rep(y_vals[2], 2), z = z_vals),
    list(x = rep(x_vals[2], 2), y = rep(y_vals[1], 2), z = z_vals),
    list(x = rep(x_vals[2], 2), y = rep(y_vals[2], 2), z = z_vals)
  )

  for (edge in lines) {
    plot <- plot %>% add_trace(
      type = 'scatter3d',
      mode = 'lines',
      x = edge$x,
      y = edge$y,
      z = edge$z,
      line = list(color = 'black', width = 2),
      showlegend = FALSE
    )
  }

  return(plot)
}

# Add the box outlines
p_with_box <- add_box_edges(p)
p_with_box



# Add contours
fig <- p_with_box %>% 
add_surface(opacity = 0.7,
  contours = list(
    z = list(
      show=TRUE,
      usecolormap=TRUE,
      highlightcolor="#ff0000",
      project=list(z=TRUE)
      )
    )
  )


fig %>% layout(scene = list(camera = list(eye = list(x = -1.25, y = 1.25, z = 1.25))))
```

```{r}
# library(rgl)
# with(random_large, {
#   plot3d(AOO_1, n_changes, j_dissim,
#          col = "blue", 
#          size = 3, 
#          type = "s", 
#          xlab = "Initial Occ.", 
#          ylab = "Changes", 
#          zlab = "Jaccard")
# })

```

```{r}
library(scatterplot3d)
scatterplot3d(random_large$AOO_1, 
              random_large$n_changes, 
              random_large$j_dissim,
              pch = 16, 
              color = "darkgreen",
              xlab = "Initial Occupancy", 
              ylab = "Num. Changes", 
              zlab = "Jaccard")

```

```{r}
#| fig.width = 40
#| fig.height = 30

library(GGally)
GGally::ggpairs(
  random_large %>% 
    select(-run, -change_type, -universe_size), 
  progress = FALSE)

```

```{r}
library(ggplot2)
ggplot(sim_df, aes(x = AOO_1, y = AOO_2, color = j_dissim)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_c() +
  facet_wrap(~ change_type) +
  theme_minimal()

sim_df %>%
  filter(change_type == "random") %>%
ggplot(
  aes(x = log_ratio, 
      y = j_dissim), 
  alpha = 0.5,
  cex = 0.1)+
  geom_point()+
  facet_wrap(~ universe_size)+
  ggthemes::theme_par()


# Mean values (faster)
sim_df %>%
  group_by(universe_size, change_type, n_changes) %>%
  summarize(
    mean_a = mean(a),
    mean_b = mean(b),
    mean_c = mean(c),
    mean_d = mean(d),
    mean_j = mean(j_dissim),
    mean_log = mean(log_ratio)) %>%
  ggplot()+
  geom_point(
    aes(
      x = log(n_changes),
      y = mean_j
    )
  ) +
  facet_wrap(~ interaction(universe_size, change_type), scales = "free_x")+
  ggthemes::theme_few()


```

```{r}
sim_df[sample(nrow(sim_df), 500), ] %>%
  ggplot()+
  geom_point(
    aes(
      x = log(n_changes),
      y = j_dissim,
      col = log(AOO_1)
    )
  ) +
  facet_wrap(~ interaction(universe_size, change_type), 
             scales = "fixed")+
  ggthemes::theme_few()
```

```{r}

sim_df %>%
  filter(universe_size == "10") %>%
  ggplot()+
  geom_point(
    aes(
      x = n_changes,
      y = j_dissim
    )
  ) +
  facet_wrap(~ change_type, 
             scales = "fixed")+
  ggthemes::theme_few()

sim_df %>%
  filter(universe_size == "100") %>%
  ggplot()+
  geom_point(
    aes(
      x = n_changes,
      y = j_dissim
    )
  ) +
  facet_wrap(~ change_type, 
             scales = "fixed")+
  ggthemes::theme_few()

sim_df %>%
  filter(universe_size == "1000") %>%
  ggplot()+
  geom_point(
    aes(
      x = n_changes,
      y = j_dissim
    )
  ) +
  facet_wrap(~ change_type, 
             scales = "fixed")+
  ggthemes::theme_few()

```

```{r}
ggplot(sim_df, aes(x = a, y = j_dissim)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  facet_wrap(~change_type) +
  theme_bw()

```

```{r}
library(mgcv)

# fit gam
gam_mod <- gam(j_dissim ~ s(a) + s(b) + s(c) + s(d) + s(universe_size, k = 3) + change_type, data = sim_df)

# plot smoothed effects
plot(gam_mod, pages = 1)

```

```{r}

# significances R2 = 0.91
summary(gam_mod)

# fitted vs. observed
plot(gam_mod$fitted.values, sim_df$j_dissim)
abline(0, 1, col = "red")

```

```{r}
install.packages("gratia")  # if not already installed
library(gratia)
draw(gam_mod)


```

```{r}

smooths_df <- smooth_estimates(gam_mod)
plot_smooths(gam_mod)  # includes faceting, labels

library(ggplot2)
plot_data <- gratia::smooth_estimates(gam_mod, smooth = "s(a)")

ggplot(plot_data, aes(x = .smooth, y = .estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = .estimate - .se, ymax = .estimate + .se), alpha = 0.3) +
  labs(x = "a (shared occupancy)", y = "Partial effect on Jaccard dissimilarity",
       title = "Smooth effect of a on j_dissim") +
  theme_minimal()

```

```{r}


ggplot(sim_df, aes(x = gam_mod$fitted.values, y = j_dissim)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  labs(x = "Predicted Jaccard dissimilarity",
       y = "Observed Jaccard dissimilarity") +
  theme_classic()



```

```{r}
library(gratia)
library(ggplot2)
library(dplyr)
library(purrr)

# Get all smooth terms (excluding factor-level effects)
smooth_names <- gratia::smooths(gam_mod)

# Get residuals and predicted values
sim_df$residuals <- residuals(gam_mod)
sim_df$fitted <- fitted(gam_mod)

# Loop through each smooth term
plots <- map(smooth_names, function(sm) {
  
  sm_df <- gratia::smooth_estimates(gam_mod, smooth = sm)
  
  # Extract the underlying variable name (e.g., s(a) → a)
  varname <- gsub("s\\((.*)\\)", "\\1", sm)
  
  # Join sim_df to get change_type and raw x values
  sm_df <- sm_df %>%
    rename(!!varname := x) %>%
    left_join(sim_df[, c(varname, "change_type")], by = varname) %>%
    distinct()
  
  ggplot(sm_df, aes_string(x = varname, y = "est")) +
    geom_line() +
    geom_ribbon(aes(ymin = est - se, ymax = est + se), alpha = 0.3) +
    facet_wrap(~change_type, scales = "free_x") +
    labs(
      x = varname,
      y = "Effect on j_dissim",
      title = paste("Smooth effect of", varname, "by change_type")
    ) +
    theme_minimal()
})

# Display all plots
plots

```

```{r}
library(ggplot2)
ggplot(sim_df, aes(x = a, y = b, fill = j)) +
  geom_tile() +
  scale_fill_viridis_c() +
  facet_wrap(~ change_type) +
  theme_minimal()

```

## Old

```{r}

# Summary stats
head(jaccard_df)
str(jaccard_df)
jaccard_df %>%
  group_by(jaccard, flip_type, length, intersect, unique_1, unique_2, ) %>%
  summarize(mean_num_changes = mean(num_changes),
            sd_num_changes = sd(num_changes),
            mean_unique_1 = mean(unique_1),
            mean_unique_2 = mean(unique_2),
            .groups = "keep")


# J ~ n_changes * universe_size * change_type * AOO_1
p1 <- jaccard_df %>%
  ggplot()+
  aes(y = jaccard, x = num_changes/length)+
  facet_wrap(vars(length, flip_type))+
  geom_point(aes(col = log(N_occu_1)))+
  geom_smooth()+
  scale_color_viridis_c(option = "viridis", direction = 1)+
  theme_linedraw()

p2 <- jaccard_df %>%
  ggplot()+
  aes(y = jaccard, x = log(num_changes))+
  facet_wrap(vars(length, flip_type))+
  geom_point(aes(col = log(N_occu_1)))+
  geom_smooth()+
  scale_color_viridis_c(option = "viridis", direction = 1)+
  theme_linedraw()

plot_grid(p1,p2, labels = "auto")

# J ~ AOO_1 * N-changes * universe_size * flip_type

# Visualization
jaccard_df %>%
  ggplot(aes(y = jaccard, x = log(N_occu_1), col = log(num_changes))) +
  geom_point(alpha = 0.3) +
  #geom_line(alpha = 0.3) +
  facet_wrap(flip_type ~ length) +
  theme_classic()

ggplot(jaccard_df) +
  aes(x = floor(log(N_occu_1)), y = jaccard, colour = log(num_changes)) +
  geom_point(size = 2.5, shape = "circle small") +
  geom_line(aes(group = log(num_changes)), linewidth = 0.75) +
  scale_color_viridis_c(option = "viridis", direction = 1) +
  #scale_x_continuous(trans = "log") +
  labs(
    x = "log(Number of initially occupied sites)",
    y = "Jaccard Similarity",
    title = "Jaccard ~ Number of initially occupied sites",
    color = "log(Number of changes)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10L),
    axis.text.x = element_text(size = 10L)
  ) +
  facet_wrap(vars(flip_type, length))


jaccard_df %>%
  ggplot(aes(y = jaccard, x = log(N_occu_1), col = num_changes, group = num_changes)) +
  geom_point(alpha = 1/3) +
  geom_line(alpha = 1/3) +
  facet_wrap(flip_type ~ length) +
  theme_linedraw()


jaccard_df %>% filter(length == 10, num_changes == 1, flip_type == "from_1_to_0") %>%
  ggplot()+
  aes(y = jaccard, x = N_occu_1) +
  geom_line()+
  geom_point()



jaccard_df %>% filter(num_changes < length-(length*0.1)) %>%
  ggplot(aes(y = jaccard, x = log(occu1), col = num_changes, group = num_changes))+
  geom_point(alpha = 1/3)+
  geom_line(alpha = 1/3)+
  facet_wrap(flip_type~length)+
  theme_classic()

ggplot(jaccard_df) +
  aes(x = floor(log(N_occu_1)), y = jaccard, colour = log(num_changes)) +
  geom_point(size = 2.5, shape = "circle small") +
  geom_line(aes(group = log(num_changes)), linewidth = 0.75) +
  scale_color_viridis_c(option = "viridis", direction = 1) +
  #scale_x_continuous(trans = "log") +
  labs(
    x = "log(Number of initially occupied sites)",
    y = "Jaccard Similarity",
    title = "Jaccard ~ Number of initially occupied sites",
    color = "log(Number of changes)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10L),
    axis.text.x = element_text(size = 10L)
  ) +
  facet_wrap(vars(flip_type, length))


jaccard_df %>%
  filter(flip_type == "random") %>% filter(num_changes < length-(length*0.1)) %>%
  group_by(jaccard, flip_type, length, intersect, unique_1, unique_2, N_occu_1, N_occu_2) %>%
  summarize(mean_num_changes = mean(num_changes),
            sd_num_changes = sd(num_changes),
            .groups = "keep") %>%
  ggplot(aes(y=jaccard, x =(N_occu_1), group = length, col = (mean_num_changes)), alpha = 0.1)+
  geom_jitter()+
  #geom_line()+
  theme_linedraw()


jaccard_df %>%
  filter(flip_type == "from_1_to_0") %>%
  group_by(jaccard, flip_type, length, intersect, unique_1, unique_2, N_occu_1, N_occu_2) %>%
  summarize(mean_num_changes = mean(num_changes),
            sd_num_changes = sd(num_changes),
            .groups = "keep") %>%
  ggplot(aes(y=jaccard, x =log(N_occu_1), group = length, col = (mean_num_changes)), alpha = 0.1)+
  geom_jitter()+
  geom_line()+
  theme_linedraw()


jaccard_df %>%
  group_by(jaccard, flip_type, length, intersect, unique_1, unique_2, N_occu_1, N_occu_2) %>%
  summarize(mean_num_changes = mean(num_changes),
            sd_num_changes = sd(num_changes),
            .groups = "keep") %>%
  ggplot(aes(y=jaccard, x =log(N_occu_1), group = length, col = (mean_num_changes)), alpha = 0.1)+
  geom_jitter()+
  #geom_line()+
  theme_linedraw()+
  facet_wrap(~flip_type)


# J ~ log(n_changes)
p_l <- lapply(unique(jaccard_df$length), function(length_i) {
  lapply(unique(jaccard_df$flip_type), function(type_i) {
    jaccard_df %>%
      filter(length == length_i, flip_type == type_i) %>%
      ggplot(aes(y = jaccard, x = log(num_changes), col = log(N_occu_1))) +
      geom_point() +
      geom_line(aes(group = log(N_occu_1)))+
      scale_color_viridis_c(option = "viridis", direction = 1) +
      labs(
        title = "Jaccard ~ log Number of changes",
        subtitle = paste0("Length = ", length_i, ", Flip Type = ", type_i)
      )
  }) %>% setNames(unique(jaccard_df$flip_type))  # Name the inner list by flip_type
}) %>% setNames(unique(jaccard_df$length))  # Name the outer list by length


plot_grid(plotlist = unlist(p_l, recursive = FALSE), ncol = 2)
ggsave("jaccard_nchanges_grid.pdf", last_plot(), device = "pdf")


jaccard_df %>% 
  mutate(log_num_changes = log(num_changes),
         log_N_occu_1 = log(N_occu_1)) %>%
  filter(flip_type == "random")%>%
  ggplot(aes(y = jaccard, x = log_num_changes, col = log_N_occu_1))+
  geom_point()+
  geom_line(aes(group = N_occu_1))+
  scale_color_viridis_c(option = "viridis", direction = 1) +
  facet_wrap(~ length, scales = "free")+
      labs(
        title = "Jaccard ~ log Number of changes")+
  theme_minimal()

jaccard_df %>%   
  mutate(log_num_changes = log(num_changes),
         log_N_occu_1 = log(N_occu_1)) %>% 
  filter(flip_type == "random") %>%
    ggplot(aes(y = jaccard, x = log_num_changes, col = factor(N_occu_1)))+
  geom_point()+
  geom_line(aes(group = interaction(N_occu_1, N_occu_2)))+
  scale_color_viridis_d(option = "viridis", direction = 1) +
  facet_wrap(~ length, scales = "free")+
      labs(
        title = "Jaccard ~ log Number of changes")+
  theme_minimal()



# List all grouping variables
group_vars <- c("length", "flip_type", "num_changes", "N_occu_1")

# Split data by unique combinations of the four variables
df_list <- split(jaccard_df, jaccard_df[group_vars])

# Apply the plotting function over each subset
p_l <- lapply(df_list, function(df_subset) {
  ggplot(df_subset, aes(y = jaccard, x = N_occu_1, col = log(num_changes))) +
    geom_point() +
    labs(
      title = "Jaccard ~ Number of Occupied Sites",
      subtitle = paste0(
        "Length = ", unique(df_subset$length), ", ",
        "Flip Type = ", unique(df_subset$flip_type), ", ",
        "Var3 = ", unique(df_subset$var3), ", ",
        "Var4 = ", unique(df_subset$var4)
      )
    )
})

plot_grid(plot_list = p_l)
```

```{r}
## Heat maps
res_proportional <- jaccard_df %>% select(-occu1, -run) %>%
  group_by(length, flip_type, num_changes) %>%
  mutate(prop_occu_1 = N_occu_1/length,
         prop_occu_2 = N_occu_2/length,
         prop_intersect = intersect/length,
         prop_union = sum(unique_1, unique_2, intersect)/length)

res_proportional


# Create a heatmap plot
ggplot(jaccard_df %>% filter(flip_type == "random"), aes(x = N_occu_2, y = N_occu_1, fill = jaccard)) +
  geom_tile(color = "white") + # Add gridlines
  scale_fill_gradient(low = "blue", high = "red", name = "Expected Jaccard") +
  labs(
    title = "Expected Jaccard Similarity",
    subtitle = "Based on Set Occupancy and Universe Size",
    x = "Set 2 Occupancy",
    y = "Set 1 Occupancy"
  ) +
  theme_minimal() +
  facet_wrap(~factor(length))+
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )


```
